### 2017-05-17 paper/content/mapSkeletons.tex
renamed \code to \inlinecode

\FloatBarrier
\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
\subsection{Parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parMap}
	\caption{Schematic depiction of parMap}
	\label{fig:parMapImg}
\end{figure}
\code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}
\caption{Definition of parMap}
\label{fig:parMap}
\end{figure}

\subsection{Lazy parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parMapStream}
	\caption{Schematic depiction of parMapStream}
	\label{fig:parMapStreamImg}
\end{figure}
As \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}. So it makes sense to also have a \code{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> arr a b -> arr [a] [b]
parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
\end{lstlisting}
\caption{Definition of parMapStream}
\label{fig:parMapStream}
\end{figure}

\subsection{Statically load-balancing parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farm}
	\caption{Schematic depiction of farm}
	\label{fig:farmImg}
\end{figure}
A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle

@@

shuffle :: (Arrow arr) => arr [[a]] [a]
shuffle = arr (concat . transpose)
\end{lstlisting}
\caption{Definition of farm. \code{unshuffle}, \code{takeEach}, \code{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
\label{fig:farm}
\end{figure}

\subsection{farmChunk}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farmChunk}
\caption{Schematic depiction of farmChunk}
\label{fig:farmChunkImg}
\end{figure}
Since a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict. So we define \code{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \code{parEvalNLazy} instead of \code{parEvalN}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
	ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
farmChunk conf chunkSize numCores f =
	unshuffle numCores >>>
	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
	shuffle
\end{lstlisting}
\caption{Definition of farmChunk}
\label{fig:farmChunk}
\end{figure}

>>>>

\FloatBarrier
\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
\subsection{Parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parMap}
	\caption{Schematic depiction of parMap}
	\label{fig:parMapImg}
\end{figure}
\inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs. We can implement it with \inlinecode{ArrowParallel} by repeating an arrow \inlinecode{arr a b} and then passing it into \inlinecode{parEvalN} to get an arrow \inlinecode{arr [a] [b]}.
Just like \inlinecode{parEvalN}, \inlinecode{parMap} is 100 \% strict.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}
\caption{Definition of parMap}
\label{fig:parMap}
\end{figure}

\subsection{Lazy parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parMapStream}
	\caption{Schematic depiction of parMapStream}
	\label{fig:parMapStreamImg}
\end{figure}
As \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \inlinecode{parEvalN} compared to \inlinecode{parEvalNLazy}. So it makes sense to also have a \inlinecode{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \inlinecode{parMap}, but uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> arr a b -> arr [a] [b]
parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
\end{lstlisting}
\caption{Definition of parMapStream}
\label{fig:parMapStream}
\end{figure}

\subsection{Statically load-balancing parallel map}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farm}
	\caption{Schematic depiction of farm}
	\label{fig:farmImg}
\end{figure}
A \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \inlinecode{ArrowParallel} we gave in this paper). This can be quite wasteful and a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \inlinecode{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle

@@

shuffle :: (Arrow arr) => arr [[a]] [a]
shuffle = arr (concat . transpose)
\end{lstlisting}
\caption{Definition of farm. \inlinecode{unshuffle}, \inlinecode{takeEach}, \inlinecode{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
\label{fig:farm}
\end{figure}

\subsection{farmChunk}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/farmChunk}
\caption{Schematic depiction of farmChunk}
\label{fig:farmChunkImg}
\end{figure}
Since a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \inlinecode{parMap} with a different work distribution, it is, again, 100\% strict. So we define \inlinecode{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
	ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
farmChunk conf chunkSize numCores f =
	unshuffle numCores >>>
	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
	shuffle
\end{lstlisting}
\caption{Definition of farmChunk}
\label{fig:farmChunk}
\end{figure}

### 2017-05-17 paper/content/motivation.tex
renamed \code to \inlinecode

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours. and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \code{map} with and without load balancing, \code{map-reduce}) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\code{pipe}, \code{ring}, \code{torus}). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

>>>>

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours. and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \inlinecode{map} with and without load balancing, \inlinecode{map-reduce}) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\inlinecode{pipe}, \inlinecode{ring}, \inlinecode{torus}). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

### 2017-05-17 paper/content/parallelHaskells.tex
renamed \code to \inlinecode

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:

\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
\caption{parEvalN's type signature}
\label{fig:parEvalNTypeSig}
\end{figure}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of parEvalN}
	\label{fig:parEvalN}
\end{figure}
\olcomment{make them to real figures? with environment, reference and such?}
Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \code{parEvalN} with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \code{using :: a -> Strategy a -> a}. For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}. We then evaluate this lazy list \code{[b]} according to a \code{Strategy [b]} with the \code{using :: a -> Strategy a -> a} operator. We construct this strategy with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \code{parEvalN} have the correct evaluation order, we annotate the computation with \code{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \code{parEvalN} computations. This is particularly necessary in circular communication topologies like in the \code{torus} or \code{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \code{pseq} during testing for this paper.

@@

\subsubsection{ParMonad}
The \code{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.

Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell. Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}. We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}. This results in \code{Par [b]}. We finally execute this process with \code{runPar} to finally get \code{[b]} again.

@@

While Eden also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
%\\
%\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
%\\
a \code{spawnF} function that
%This 
allows us to define \code{parEvalN} directly:

>>>>

In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel:

\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
\caption{parEvalN's type signature}
\label{fig:parEvalNTypeSig}
\end{figure}
\begin{figure}[h]
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of parEvalN}
	\label{fig:parEvalN}
\end{figure}
\olcomment{make them to real figures? with environment, reference and such?}
Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \inlinecode{using :: a -> Strategy a -> a}. For \inlinecode{parEvalN} this means that we can just apply the list of functions \inlinecode{[a -> b]} to the list of inputs \inlinecode{[a]} by zipping them with the application operator \inlinecode{\$}. We then evaluate this lazy list \inlinecode{[b]} according to a \inlinecode{Strategy [b]} with the \inlinecode{using :: a -> Strategy a -> a} operator. We construct this strategy with \inlinecode{parList :: Strategy a -> Strategy [a]} and \inlinecode{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form. To ensure that programs that use \inlinecode{parEvalN} have the correct evaluation order, we annotate the computation with \inlinecode{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \inlinecode{parEvalN} computations. This is particularly necessary in circular communication topologies like in the \inlinecode{torus} or \inlinecode{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \inlinecode{pseq} during testing for this paper.

@@

\subsubsection{ParMonad}
The \inlinecode{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.

Our parallel evaluation function \inlinecode{parEvalN} can be defined by zipping the list of \inlinecode{[a -> b]} with the list of inputs \inlinecode{[a]} with the application operator \inlinecode{\$} just like with Multicore Haskell. Then, we map over this not yet evaluated lazy list of results \inlinecode{[b]} with \inlinecode{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \inlinecode{[Par (IVar b)]}, which we convert to \inlinecode{Par [IVar b]} with \inlinecode{sequenceA}. We wait for the computations to finish by mapping over the \inlinecode{IVar b}'s inside the \inlinecode{Par} monad with \inlinecode{get}. This results in \inlinecode{Par [b]}. We finally execute this process with \inlinecode{runPar} to finally get \inlinecode{[b]} again.

@@

While Eden also comes with a monad \inlinecode{PA} for parallel evaluation, it also ships with a completely functional interface that includes
%\\
%\inlinecode{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
%\\
a \inlinecode{spawnF} function that
%This 
allows us to define \inlinecode{parEvalN} directly:

### 2017-05-17 paper/content/syntacticSugar.tex
renamed \code to \inlinecode

\subsection{Syntactic Sugar} \label{syntacticSugar}
For basic arrows, we have the \code{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once. This can easily be translated into a parallel version \code{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in Fig.~\ref{fig:|***|}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{lstlisting}
\caption{Definition of (|***|) - the parallel version of (***)}
\label{fig:|***|}
\end{figure}
% With this we can analogously to the serial \code{&&&}
We define the parallel \code{(|\&\&\&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{lstlisting} % $ %% formatting
\caption{Definition of (|\&\&\&|) - the parallel version of (\&\&\&)}
\label{fig:|&&&|}
\end{figure}

>>>>

\subsubsection{Syntactic Sugar} \label{syntacticSugar}
For basic arrows, we have the \inlinecode{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into an arrow \inlinecode{arr (a, c) (b, d)} which does both computations at once. This can easily be translated into a parallel version \inlinecode{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:|***|}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{lstlisting}
\caption{Definition of (|***|) - the parallel version of (***)}
\label{fig:|***|}
\end{figure}
% With this we can analogously to the serial \inlinecode{&&&}
We define the parallel \inlinecode{(|\&\&\&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \inlinecode{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{lstlisting} % $ %% formatting
\caption{Definition of (|\&\&\&|) - the parallel version of (\&\&\&)}
\label{fig:|&&&|}
\end{figure}
